---
title: "iChip 1 - 6 European Samples Prepare for Imputation Submission"
author: "Translational Genomics Group"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(dplyr)
```

## Prepare VCF for Imputation Submission
### Download HRC Reference 
This file has already been downloaded to a shared directory. 
```{bash eval = FALSE}
wget -p /mnt/share6/SHARED_DATASETS/Haplotype_Reference_Consortium ftp://ngs.sanger.ac.uk/production/hrc/HRC.r1-1/HRC.r1-1.GRCh37.wgs.mac5.sites.tab.gz &>/dev/null
```

```{bash eval = FALSE}
gunzip -k /mnt/share6/SHARED_DATASETS/Haplotype_Reference_Consortium/HRC.r1-1.GRCh37.wgs.mac5.sites.tab.gz &>/dev/null
```

### Download HRC Check Tool 
```{bash eval = FALSE}
wget http://www.well.ox.ac.uk/~wrayner/tools/HRC-1000G-check-bim-v4.2.9.zip -O temp.zip; unzip temp.zip; rm temp.zip
```

### Download Will Rayner Strand FIle to Align to Reference
Download Immuno_BeadChip_11419691_B Strand Files for iChip 1 to 6 and Ref/Alt File
```{bash eval = FALSE, message = FALSE}
wget http://www.well.ox.ac.uk/~wrayner/strand/Immuno_BeadChip_11419691_B-b37-strand.zip -O temp.zip; unzip temp.zip; rm temp.zip &>/dev/null

wget http://www.well.ox.ac.uk/~wrayner/strand/RefAlt/Immuno_BeadChip_11419691_B-b37.strand.RefAlt.zip -O temp.zip; unzip temp.zip; rm temp.zip &>/dev/null
```

### B. Create List of SNPs to Flip
```{r}
strand <- read_tsv("immuno_beadchip_11419691_b-b37.strand", 
                   col_names = c("SNP", "Chr", "BP", "Match", "Strand", "Allele"))

strand %>%
  filter(Strand == "-") %>%
  select(SNP) %>%
  write_tsv("temp_flip_ichip1t6.tsv", col_names = FALSE)
```

### C. Flip SNPs to Convert TOP to Forward/Ref in Plink
```{bash}
plink \
--bfile ../3.qc/qc_eur_cohort_split_ichip1t6 \
--flip temp_flip_ichip1t6.tsv \
--make-bed \
--out temp_ichip1t6
```

### D. Set Ref/Alt Allele 
```{bash}
plink \
--bfile temp_ichip1t6 \
--reference-allele Immuno_BeadChip_11419691_B-b37.strand.RefAlt 2 \
--chr 1-22 \
--make-bed \
--out temp1_ichip1t6
```

### E. Calcualte Allele Frequency
```{bash}
plink \
--bfile temp1_ichip1t6 \
--keep-allele-order \
--freq \
--out temp1_ichip1t6
```

### F. Final Checks
```{bash eval = FALSE}
perl HRC-1000G-check-bim.pl \
-b temp1_ichip1t6.bim \
-f temp1_ichip1t6.frq \
-r /mnt/share6/SHARED_DATASETS/Haplotype_Reference_Consortium/HRC.r1-1.GRCh37.wgs.mac5.sites.tab \
-h
```

- 4 markers, match by name to HRC but have a different position (same CHR), will exclude. rs1377587 rs1319548 rs9258651	rs2085508
- 66 markers are on the wrong strand when compared to HRC reference after strand flips from Will's files. Will re-flip possible difference references used. 
- 854 with wrong ref assignemnt

### Remove 4 SNPs with different positions
```{bash}
plink \
--bfile temp1_ichip1t6 \
--exclude Position-temp1_ichip1t6-HRC.txt \
--make-bed \
--out temp2_ichip1t6
```

### Exclude SNPS
```{bash}
plink \
--bfile temp2_ichip1t6 \
--exclude Exclude-temp1_ichip1t6-HRC.txt \
--make-bed \
--out temp3_ichip1t6
```

### Flip 
```{bash}
plink \
--bfile temp3_ichip1t6 \
--flip Strand-Flip-temp1_ichip1t6-HRC.txt \
--make-bed \
--out temp4_ichip1t6
```

### Set Ref/Alt
```{bash}
plink \
--bfile temp4_ichip1t6 \
--reference-allele Force-Allele1-temp1_ichip1t6-HRC.txt \
--make-bed \
--out ichip1t6_eur_qc_pre_impute
```

```{r}
freq_plot <- read_tsv("FreqPlot-temp1_ichip1t6-HRC.txt", col_names = FALSE)

freq_plot %>%
  ggplot(aes(x = X2, y = X3)) +
  geom_point(alpha = 1/500)

```

## Convert to VCF using Plink
Using Plink1.9 becuase plink2 outputs VCFv4.3 which is not yet supported by the server.
Plink1.9 on vcf recode defualts to A2 allele as reference, so forcing a2 allele to curent a1. 

```{bash eval = FALSE}
for i in {1..22}; do
    plink --bfile ichip1t6_eur_qc_pre_impute \
    --chr ${i} \
    --real-ref-alleles \
    --recode vcf \
    --out ichip1t6_eur_vcfs/ichip1t6_eur_chr${i}
done
```

## Create Sorted and Compressed VCF using VCFtools and tabix (including bgzip)
```{bash eval = FALSE}
for i in {1..22}; do
    vcf-sort ichip1t6_eur_vcfs/ichip1t6_eur_chr${i}.vcf | bgzip -c > ichip1t6_eur_vcfs/ichip1t6_eur_chr${i}.vcf.gz
done
```

```{r}
file.remove(list.files(pattern = "^temp", full.names = TRUE))
```
Submit to Michigan Imputation Server. 
